Understanding the `aghq` package

### 2022-10-04 src/explore_aghq/aghq.Rmd
Work on understanding aghq vignette

Use the `aghq` package with `k = 3`:

@@

```{r}
summary(quad)
plot(quad)
```

The `quad` object contains:

```{r}
str(quad)
```

The nodes and weights of the posterior are:

@@

The logarithm of the normalising constant (should correspond to `logpost - logpost_normalized`):

@@

```{r}
plot <- data.frame(x = seq(from = -10, to = 10, by = 0.1)) %>%
  mutate(
    y = f(x)
  ) %>%
  ggplot(aes(x = x, y = y)) +
    geom_line() +
    theme_minimal()

@@

#' The scale for weights is a bit silly
plot +
  geom_point(
    data = data.frame(nodes = gg$nodes, weights = gg$weights, y = 0), 
    aes(x = nodes, y = y, size = weights),
    col = cbpalette[2]
  )
```

@@

Or more manually via:

```{r}
sum(f(gg$nodes) * gg$weights)

#' Note: you can also use these functions to get the nodes and weights
# mvQuad::getNodes(gg)
# mvQuad::getWeights(gg)
```

This is surprisingly close to the real answer, which is:

@@

What about in two dimensions?

@@

@@

Equivalent to:

@@

```{r}
plot2 +
  geom_point(
    data = mvQuad::getNodes(gg2) %>%
            as.data.frame() %>%
            mutate(weights = mvQuad::getWeights(gg2)), 
    aes(x = V1, y = V2, size = weights, fill = 1),
    col = cbpalette[2]
  )
```

Do the quadrature using:

```{r}
#' Need f to be defined like this for some reason
f <- function(x) {
  apply(x, 1, function(y) prod(exp(-(y^2) / 2)))  
}

@@

As compare with the true answer:

@@

```{r}
# Gauss Hermite Quadrature
# Showing that the nodes are zeros of Hermite polynomial
Hk <- as.function(mpoly::hermite(3, kind = "he"))
nn <- mvQuad::getNodes(gg)
round(Hk(nn))

# Adaptive: scale the nodes and weights by the mode and curvature
theta_hat <- c(2, 3)
theta_hess <- matrix(c(3, 1, 1, 5), nrow = 2, ncol = 2)
theta_hess_inv <- solve(theta_hess)
k <- 3

# Can be rescaled using mvQuad via
int_grid_software <- mvQuad::createNIGrid(2, "GHe", k)
mvQuad::rescale(int_grid_software, m = theta_hat, C = theta_hess_inv, dec.type = 2)
# or manually via
int_grid_manual <- mvQuad::createNIGrid(2, "GHe", k)
nn_rescaled <- mvQuad::getNodes(int_grid_manual) %*% chol(theta_hess_inv)
nn_rescaled[, 1] <- nn_rescaled[, 1] + theta_hat[1]
nn_rescaled[, 2] <- nn_rescaled[, 2] + theta_hat[2]
# and for the weights
mvQuad::getWeights(int_grid_software)
ww_rescaled <- mvQuad::getWeights(int_grid_manual) * det(chol(theta_hess_inv))

# k = 1 then ACHQ is a Laplace approximation
```

@@

```{r}
library(TMB)
data("tswv", package = "EpiILMCT")

precompile()
compile("disease.cpp")
dyn.load(dynlib("disease"))

dat <- tswv$tswvsir
dat$epidat <- dat$epidat[order(dat$epidat[, 4]), ]
I <- dat$epidat[, 4]
R <- dat$epidat[, 2]
infected <- !is.infinite(I)

datlist <- list(
  D = as.matrix(dist(dat$location[dat$epidat[, 1], ])),
  I = I,
  R = R,
  infected = as.numeric(infected[infected])
)

ff <- TMB::MakeADFun(
  data = datlist,
  parameters = list(theta1 = 0, theta2 = 0),
  DLL = "disease",
  ADreport = FALSE,
  silent = TRUE
)

tm <- Sys.time()
# c(0, 0) is start and k = 7, ilm = individual-level model (?)
tmp <- capture.output(
  ilmquadrature <- aghq::aghq(ff, 7, c(0, 0))
)
runtime <- difftime(Sys.time(), tm, units = "secs")

summary(ilmquadrature)

# compute_moment is the mean (only?)
ilm_means <- aghq::compute_moment(ilmquadrature$normalized_posterior, function(x) exp(x))
names(ilm_means) <- c("alpha", "beta")

# q is a vector of numeric values in (0, 1) giving the quantiles you want
my_q <- c(0.025, 0.1, 0.5, 0.9, 0.975)
exp(aghq::compute_quantiles(ilmquadrature$marginals[[1]], q = my_q)) # alpha
exp(aghq::compute_quantiles(ilmquadrature$marginals[[2]], q = my_q)) # beta
```

>>>>

Use the `aghq` package with `k = 3`.
Note that all `aghq::aghq::` is doing is normalising the posterior.
We have already provided the exact posterior to it via `fn$ff`.

@@

```{r}
summary(quad)
```

We can plot the log marginal posterior for `theta1` in its normalised and unnormalised versions.
The unnormalised version doesn't require anything from `aghq`.

```{r}
data.frame(
  theta1 = seq(from = 1, to = 2, by = 0.01)
) %>%
  mutate(
    f_unnormalised = ff$fn(theta1),
    f_normalised = f_unnormalised - quad$normalized_posterior$lognormconst
  ) %>%
  pivot_longer(
    cols = starts_with("f"),
    names_sep = "_",
    names_to = c("function", "normalised")
  ) %>%
  ggplot(aes(x = theta1, y = value)) +
    geom_line() +
    facet_grid(~normalised) +
    labs(x = "theta1", y = "log(posterior)") +
    theme_minimal()
```

The contents of the `quad` object can be viewed using `str(quad)`.
One part of the contents are the nodes and weights used for quadrature, which are:

@@

The logarithm of the normalising constant (should correspond to `logpost - logpost_normalised`, the difference between the two curves above):

@@

```{r}
plot <- data.frame(x = seq(from = -10, to = 10, by = 0.1)) %>%
  mutate(
    y = f(x)
  ) %>%
  ggplot(aes(x = x, y = y)) +
    geom_line() +
    labs(x = "x", y = "f(x)") +
    theme_minimal()

@@

plot +
  geom_point(
    data = data.frame(nodes = gg$nodes, weights = gg$weights, y = 0), 
    aes(x = nodes, y = y, size = weights),
    col = cbpalette[2]
  ) +
  scale_size_continuous(limits = c(1, 2))
```

@@

This isn't doing anything fancy, it's just multiplying the values at the nodes by the weights:

```{r}
sum(f(gg$nodes) * gg$weights)
```

Note that you can also use these functions to get the nodes and weights: `mvQuad::getNodes(gg)`, `mvQuad::getWeights(gg)`.
The answer we get with `aghq` is exact in this case, as compared with the true answer of $\sqrt {2 \pi}$ because the target function is a log-quadratic:

@@

What about a function in two dimensions?
Take this one, where each marginal is the same as above:

@@

Let's create a quadrature grid again:

@@

This is equivalent to what can be achieved using the `gg` object from the one-dimensional example:

@@

What do these nodes and weights look like on the plot?

```{r}
plot2 +
  geom_point(
    data = mvQuad::getNodes(gg2) %>%
            as.data.frame() %>%
            mutate(weights = mvQuad::getWeights(gg2)), 
    aes(x = V1, y = V2, size = weights, fill = 1),
    col = cbpalette[2]
  ) +
  labs(col = "f(x1, x2)") +
  scale_size_continuous(limits = c(2, 4))
```

Ok! So now let's use `mvQuad` to do the quadrature:

```{r}
#' Need f to be defined like this for some reason, can probably be fixed
f <- function(x) {
  apply(x, 1, function(y) prod(exp(-(y^2) / 2)))  
}

@@

Again, in this case the quadrature is exact

@@

See Definition 1 of the [theory paper](https://arxiv.org/pdf/2102.06801.pdf) for more on when the integration will be exact.

## Gauss-Hermite Quadrature

Define the Hermite polynomial with $k = 3$ as:

```{r}
Hk <- as.function(mpoly::hermite(3, kind = "he"))

plot <- data.frame(x = seq(-3, 3, by = 0.1)) %>%
  mutate(Hkx = Hk(x)) %>%
  ggplot(aes(x = x, y = Hkx)) +
    geom_line() + 
    labs(x = "x", y = "H3(x)") +
    theme_minimal()

In Gauss-Hermite quadrature, we pick nodes which are zeroes of the Hermite polynomial:

```{r}
nn <- mvQuad::getNodes(gg)

plot +
  geom_point(
    data = data.frame(x = nn, y = Hk(nn)), aes(x = x, y = y),
    col = cbpalette[2], size = 3
  )
```

Note that these nodes are chosen *only* based on the Hermite polynomial.
The function to be integrated has not played any role in their choice!
This is probably not a good idea.
So in adaptive quadrature, the function plays some role in how the points are chosen.
Specifically in adaptive GHQ it's done by scaling the nodes and weights by the mode and curvature at the mode of the function to be integrated.
Imagine we have some fake values for the mode `theta_hat` and curvature at the mode `theta_hess` as follows:

```{r}
theta_hat <- c(2, 3)
theta_hess <- matrix(c(3, 1, 1, 5), nrow = 2, ncol = 2)
theta_hess_inv <- solve(theta_hess)
```

Then the rescaling can be done with `mvQuad` easily via:

```{r}
k <- 3
int_grid_software <- mvQuad::createNIGrid(2, "GHe", k)
mvQuad::rescale(int_grid_software, m = theta_hat, C = theta_hess_inv, dec.type = 2)
```

All that is doing manually to the nodes is:

```{r}
int_grid_manual <- mvQuad::createNIGrid(2, "GHe", k)
nn_rescaled <- mvQuad::getNodes(int_grid_manual) %*% chol(theta_hess_inv)
nn_rescaled[, 1] <- nn_rescaled[, 1] + theta_hat[1]
nn_rescaled[, 2] <- nn_rescaled[, 2] + theta_hat[2]
```

Or to the weights is:

```{r}
mvQuad::getWeights(int_grid_software)
ww_rescaled <- mvQuad::getWeights(int_grid_manual) * det(chol(theta_hess_inv))
```

@@

Ok, let's now move on to an example with real data.

```{r}
data("tswv", package = "EpiILMCT")
dat <- tswv$tswvsir
dat$epidat <- dat$epidat[order(dat$epidat[, 4]), ]
I <- dat$epidat[, 4]
R <- dat$epidat[, 2]
infected <- !is.infinite(I)

datlist <- list(
  D = as.matrix(dist(dat$location[dat$epidat[, 1], ])),
  I = I,
  R = R,
  infected = as.numeric(infected[infected])
)
```

```{r}
precompile()
compile("disease.cpp")
dyn.load(dynlib("disease"))
```

Create the unnormalised negative log posterior using `TMB::MakeADFun` as follows:

```{r}
ff <- TMB::MakeADFun(
  data = datlist,
  parameters = list(theta1 = 0, theta2 = 0),
  DLL = "disease",
  ADreport = FALSE,
  silent = TRUE
)
```

Perform AGHQ with $k = 7$, starting optimisation from $(0, 0)$:

```{r}
tm <- Sys.time()

tmp <- capture.output(
  ilm_quadrature <- aghq::aghq(ff, 7, c(0, 0))
)

runtime <- difftime(Sys.time(), tm, units = "secs")
```

The time this took was:

```{r}
runtime
```

Note that as there are $k = 7$ quadrature points per dimension, in total there are $7^2 = 49$:

```{r}
summary(ilm_quadrature)
```
The parameters $\alpha = \exp(\theta_1)$ and $\beta = \exp(\theta_2)$, which we can compute summaries for as follows:

```{r}
ilm_means <- aghq::compute_moment(ilm_quadrature$normalized_posterior, function(x) exp(x))
names(ilm_means) <- c("alpha", "beta")
ilm_means
```

You can calculate other moments using the `ff`, `gg` or `nn` arguments to `aghq::compute_moment`.
There is also some nuance to the argument `method`, either `reuse` or `correct`.
`reuse` uses the previously computed AGHQ nodes and weights.
`correct` recomputes the mode and curvature for the product of $g$ and the posterior; it has convergence proofs, but is a bit slower and can be unstable.

We can also compute quantiles using the function `aghq::compute_quantiles`, where `q` is a vector of numeric values in $(0, 1)$.

```{r}
my_q <- c(0.025, 0.1, 0.5, 0.9, 0.975)
```

For $\alpha$:

```{r}
exp(aghq::compute_quantiles(ilm_quadrature$marginals[[1]], q = my_q))
```

For $\beta$:

```{r}
exp(aghq::compute_quantiles(ilm_quadrature$marginals[[2]], q = my_q)) # beta
```
